---
title: "Reproducible Research: Peer Assessment 1"

output: 

  html_document:

    keep_md: true
---


### 1. Loading and Preprocessing the data 
 
##### **Load the data**

```{r unzip, cache=TRUE}

     unzip("repdata_data_activity.zip",exdir = "data")

```

##### **Process/Transform the data**

###### Read the data into a data frame
```{r reading, cache=TRUE}

     rawData <- read.csv("data/activity.csv", stringsAsFactors=FALSE)
     summary(rawData)

```


###### Transform Data: convert date to POSIXct class using lubridate package

```{r date_conversion, message=FALSE}

    library(lubridate)
    rawData$date <- ymd(rawData$date)
    activity <- rawData[complete.cases(rawData),]
    str(activity)

```




### 2. What is mean total number of steps taken per day?


##### **Calculate the total number of steps taken per day**

```{r daily_total,message=FALSE}

library(dplyr)

StepsPerDay<-activity%>%group_by(activity$date)%>%summarise(TotSteps=sum(steps))

```

##### **Histogram for total number of steps taken per day**

```{r barplot,fig.width=12}

barplot(height = StepsPerDay$TotSteps,names.arg=StepsPerDay$date,cex.names=0.68,las=3,col="orange",main = "Histogram for Total Steps Per Day")

abline(h=median(StepsPerDay$TotSteps), lty=2,lwd=3, col="black")

abline(h=mean(StepsPerDay$TotSteps), lty=2,lwd=3, col="red")

text(x = 0,y=median(StepsPerDay$TotSteps),pos=3,labels = "median")

text(x = 0,y=mean(StepsPerDay$TotSteps),pos=1,labels = "mean",col="red")

```



##### **Calculate and report the MEAN and MEDIAN of the total number of steps taken per day**



```{r summary}

mean_steps <- mean(StepsPerDay$TotSteps,na.rm=TRUE)

median_steps <- median(StepsPerDay$TotSteps,na.rm=TRUE)

```

Mean and median of the total number of steps taken per day are `r mean_steps` steps and `r median_steps` steps, respectively.



### 3.What is the average daily activity pattern?

##### **Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**



```{r daily, fig.width=12}

library(dplyr,quietly = TRUE)

daily_activity_patterns <- activity %>% group_by(interval) %>% summarise(average=mean(steps,na.rm=TRUE))

plot(x = 1:nrow(daily_activity_patterns),y = daily_activity_patterns$average,type = "l",

     col = "red", xaxt = "n",xlab="Intervals", 

     ylab = "Average number of steps taken", main = "Time series plot of avg. steps taken in 5 min. intervals")

axis(1,labels=daily_activity_patterns$interval[seq(1,288,12)],

     at = seq_along(daily_activity_patterns$interval)[seq(1,288,12)])

```



##### **Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r daily2}

max_numb_steps_interval <- filter(daily_activity_patterns,average==max(average))

```



Interval **"`r max_numb_steps_interval$interval`"** contains on average the maximum number of steps (**`r round(max_numb_steps_interval$average,2)` steps**).



### 4.Imputing missing values



##### **Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**



```{r missing}

na_number <- sum(is.na(activity$steps))

na_number

percentage_na <- mean(is.na(activity$steps))

percentage_na

```

Total number of missing values in the dataset amounts to **`r na_number` ** (what is **`r round(percentage_na*100,1)`** % of total observations).



##### **Devise a strategy for filling in all of the missing values in the dataset**



As the number of missing values in this dataset is fairly large, we cannot be sure if there is no bias introduced by missing values. Therefore we impute missing values based on average number of steps in particular 5-minutes interval. 



##### **Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r na_imputing, cache=TRUE}

non_NAs <- numeric(nrow(activity))

for (i in 1:nrow(activity))
{
      if (is.na(activity[i,"steps"])==TRUE)
          {
                  non_NAs[i]<-filter(daily_activity_patterns,interval==activity[i,"interval"]) %>% select(average)
            } 
        else
            {
                  non_NAs[i]<-activity[i,"steps"]

            }

}

activity_non_NAs<-mutate(activity,steps_no_NAs=non_NAs)
head(activity_non_NAs)

```



Below code is just to verify if process of imputing missing values correctly preserved original values (lines with no NAs)   

```{r test_NAs}

check <- filter(activity_non_NAs,!is.na(steps)) %>% mutate(ok = (steps==steps_no_NAs))

mean(check$ok)



```



##### **Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day**



```{r histogram_no_NAs}

StepsPerDay_noNAs <- activity_non_NAs %>% mutate(steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(date) %>% summarise(TotSteps=sum(steps_no_NAs))

hist(StepsPerDay_noNAs$TotSteps,col="green",breaks=20,main="Histogram of total number of steps taken each day (Imputed values)",xlab="Steps per day")

abline(v=median(StepsPerDay$TotSteps),lty=3, lwd=2, col="blue")

legend(legend="median","topright",lty=3,lwd=2,bty = "n")

```



```{r summary_no_NAs}



summary(StepsPerDay_noNAs$TotSteps)

```



Imputing missing values, mean of the total number of steps taken per day  increased while median decreased,compared to estimates from the first part (ingoring missing values). Imputing missing data resulted in increase of total daily number of steps (instead of each NAs we have average that is always >=0)



### 5. Are there differences in activity patterns between weekdays and weekends?



##### **Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day**



```{r weekday}

library(lubridate)

is_weekday <-function(date){
        if(wday(date)%in%c(1,7)) result<-"weekend"
        else
                result<-"weekday"
        result
}

activity_non_NAs <- mutate(activity_non_NAs,date=ymd(date)) %>% mutate(day=sapply(date,is_weekday))

table(activity_non_NAs$day)

```


##### **Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)**


```{r weekend_comparison, cache=TRUE, fig.width=10}

library(ggplot2)

daily_activity_patterns <- activity_non_NAs %>% mutate(day=factor(day,levels=c("weekend","weekday")),steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(interval,day) %>% summarise(average=mean(steps_no_NAs))

qplot(interval,average,data=daily_activity_patterns,geom="line",facets=day~.)

```